---
title: "Proxy models + demo/geo"
format: html
editor: visual
---

## TFG: Proxy variables and models

### Data setup & cleaning

```{r}

library(tidyverse)
library(janitor)
library(skimr)
library(naniar)
library(gt)
library(patchwork)
library(corrplot)
library(MASS)
library(broom)
library(rsample)
library(yardstick)
library(mgcv)
library(purrr)
library(caret)
library(ranger)
library(doParallel)
library(nnet)

set.seed(123)

raw_data <- read_delim(
  "THE-subpilars-geodemo-data.csv",
  delim = ";",
  col_types = cols(.default = col_character()),
  show_col_types = FALSE,
  trim_ws = TRUE
)

df <- raw_data %>%
  clean_names() %>%
  mutate(across(everything(), ~ na_if(str_trim(.x), "")))

glimpse(df)
dim(df)
names(df)

to_numeric_clean <- function(x) {
  x %>%
    str_replace_all(",", ".") %>%
    as.numeric()
}

df <- df %>%
  mutate(
    world_ranking_the = as.integer(world_ranking_the),
    
    across(
      c(
        # Original subpillar / proxy variables
        total_students_enrolled,
        student_to_staff_ratio,
        estimated_staff,
        international_students_fraction,
        total_international_students,
        female_students_fraction,
        total_citations,
        citations_per_faculty,
        total_publications,
        h_index,
        x2_year_mean_citations_per_publication,
        x5_year_publication_growth_percent,
        x5_year_citation_growth_percent,
        research_productivity,
        international_co_authorship,
        alumni_impact_index_0_1000,
        
        # New demographic / geographic variables
        country_gdp_billion_usd,
        country_gdp_per_capita_usd,
        number_of_associated_institutions,
        number_of_child_institutions,
        university_age
      ),
      to_numeric_clean
    ),
    
    carbon_neutrality_net_zero_target_0_1 =
      as.integer(to_numeric_clean(carbon_neutrality_net_zero_target_0_1)),
    
    native_english_speaking_country =
      as.integer(to_numeric_clean(native_english_speaking_country)),
    
    is_private_university =
      as.integer(to_numeric_clean(is_private_university))
  )

missing_summary <- df %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_count"
  ) %>%
  mutate(
    missing_pct = round(missing_count / nrow(df) * 100, 2)
  ) %>%
  arrange(desc(missing_count))

missing_summary %>%
  filter(missing_count > 0) %>%
  gt()

df_analysis <- df %>%
  mutate(
    carbon_neutrality_net_zero_target_0_1 = factor(
      carbon_neutrality_net_zero_target_0_1,
      levels = c(0, 1),
      labels = c("No target", "Net-zero target"),
      ordered = TRUE
    ),
    
    native_english_speaking_country = factor(
      native_english_speaking_country,
      levels = c(0, 1),
      labels = c("Non-native English-speaking", "Native English-speaking")
    ),
    
    is_private_university = factor(
      is_private_university,
      levels = c(0, 1),
      labels = c("Public", "Private")
    ),
    
    country = factor(country),
    primary_research_topic = factor(primary_research_topic),
    top_collaborating_country = factor(top_collaborating_country)
  )

df_analysis_clean <- df_analysis %>%
  drop_na()

dim(df_analysis)
dim(df_analysis_clean)

df_analysis_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_count"
  ) %>%
  arrange(desc(missing_count)) %>%
  gt()

num_vars <- df_analysis_clean %>%
  dplyr::select(where(is.numeric)) %>%
  names()

cat_vars <- df_analysis_clean %>%
  dplyr::select(where(is.factor)) %>%
  names()

num_vars
cat_vars

var_labels <- c(
  # Outcome
  world_ranking_the = "THE world ranking position",
  
  # Institution identifier
  institution_name = "Institution name",
  
  # Size / institutional structure
  total_students_enrolled = "Total students enrolled",
  student_to_staff_ratio = "Student-to-staff ratio",
  estimated_staff = "Estimated staff",
  
  # Internationalisation / gender
  international_students_fraction = "International students fraction",
  total_international_students = "Total international students",
  female_students_fraction = "Female students fraction",
  
  # Research output / impact
  total_citations = "Total citations",
  citations_per_faculty = "Citations per faculty",
  total_publications = "Total publications",
  h_index = "H-index",
  x2_year_mean_citations_per_publication = "2-year mean citations per publication",
  x5_year_publication_growth_percent = "5-year publication growth (%)",
  x5_year_citation_growth_percent = "5-year citation growth (%)",
  research_productivity = "Research productivity",
  international_co_authorship = "International co-authorship",
  alumni_impact_index_0_1000 = "Alumni impact index",
  
  # Sustainability
  carbon_neutrality_net_zero_target_0_1 = "Carbon neutrality / Net-zero target",
  
  # Geographic / demographic variables
  country = "Country",
  primary_research_topic = "Primary research topic",
  country_gdp_billion_usd = "Country GDP",
  country_gdp_per_capita_usd = "Country GDP per capita",
  number_of_associated_institutions = "Number of associated institutions",
  number_of_child_institutions = "Number of child institutions",
  top_collaborating_country = "Top collaborating country",
  university_age = "University age",
  native_english_speaking_country = "Native English-speaking country",
  is_private_university = "Private university"
)

get_var_label <- function(var) {
  if (var %in% names(var_labels)) var_labels[[var]] else var
}
```

### General Distribution Analysis

```{r}

num_vars <- df_analysis_clean %>%
  dplyr::select(where(is.numeric)) %>%
  names()

cat_vars <- df_analysis_clean %>%
  dplyr::select(where(is.factor)) %>%
  names()

num_vars
cat_vars
```

### Numerical Variables

```{r}

num_summary <- df_analysis_clean %>%
  dplyr::select(all_of(num_vars)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  group_by(variable) %>%
  summarise(
    variable_label = get_var_label(first(variable)),
    n = sum(!is.na(value)),
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    min = min(value, na.rm = TRUE),
    q1 = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    q3 = quantile(value, 0.75, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    .groups = "drop"
  )

num_summary %>%
  gt()

plot_num_var <- function(data, var, bins = 30) {
  var_label <- get_var_label(var)
  
  p1 <- ggplot(data, aes(x = .data[[var]])) +
    geom_histogram(
      bins = bins,
      fill = "#0072B2",
      color = "white",
      na.rm = TRUE
    ) +
    labs(
      title = paste("Distribution of", var_label),
      x = var_label,
      y = "Count"
    ) +
    theme_minimal()
  
  p2 <- ggplot(data, aes(y = .data[[var]])) +
    geom_boxplot(
      fill = "#0072B2",
      alpha = 0.7,
      na.rm = TRUE
    ) +
    labs(
      title = paste("Boxplot of", var_label),
      x = NULL,
      y = var_label
    ) +
    theme_minimal()
  
  list(
    histogram = p1,
    boxplot = p2,
    combined = p1 / p2
  )
}

num_plot_list <- set_names(
  map(num_vars, ~ plot_num_var(df_analysis_clean, .x)),
  num_vars
)

walk(num_plot_list, ~ print(.x$combined))
```

### Categorical / Binary variables

```{r}

cat_summary <- df_analysis_clean %>%
  dplyr::select(all_of(cat_vars)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "category",
    values_transform = list(category = as.character)
  ) %>%
  count(variable, category, sort = TRUE) %>%
  group_by(variable) %>%
  mutate(
    pct = round(n / sum(n) * 100, 2),
    variable_label = get_var_label(first(variable))
  ) %>%
  ungroup()

cat_summary %>%
  gt()

plot_cat_var <- function(data, var, top_n = 15) {
  var_label <- get_var_label(var)
  
  data %>%
    mutate(
      category_plot = fct_lump_n(.data[[var]], n = top_n)
    ) %>%
    count(category_plot, sort = TRUE) %>%
    mutate(
      category_plot = fct_reorder(category_plot, n)
    ) %>%
    ggplot(aes(x = category_plot, y = n)) +
    geom_col(fill = "#0072B2") +
    coord_flip() +
    labs(
      title = paste("Distribution of", var_label),
      subtitle = paste("Top", top_n, "categories; remaining grouped as Other"),
      x = var_label,
      y = "Count"
    ) +
    theme_minimal()
}

cat_plot_list <- set_names(
  map(cat_vars, ~ plot_cat_var(df_analysis_clean, .x)),
  cat_vars
)

walk(cat_plot_list, print)
```

### Outliers analysis:

```{r}

detect_outliers_iqr <- function(x, multiplier = 3) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  
  lower <- q1 - multiplier * 5 * iqr
  upper <- q3 + multiplier * 5 * iqr
  
  x < lower | x > upper
}

outlier_summary <- map_dfr(
  num_vars,
  \(v) {
    outliers <- detect_outliers_iqr(df_analysis_clean[[v]], multiplier = 3)
    
    tibble(
      variable = v,
      variable_label = get_var_label(v),
      outlier_count = sum(outliers, na.rm = TRUE),
      outlier_pct = round(outlier_count / nrow(df_analysis_clean) * 100, 2),
      min = min(df_analysis_clean[[v]], na.rm = TRUE),
      q1 = quantile(df_analysis_clean[[v]], 0.25, na.rm = TRUE),
      median = median(df_analysis_clean[[v]], na.rm = TRUE),
      q3 = quantile(df_analysis_clean[[v]], 0.75, na.rm = TRUE),
      max = max(df_analysis_clean[[v]], na.rm = TRUE)
    )
  }
) %>%
  arrange(desc(outlier_count))

outlier_summary %>%
  gt()

outlier_vars <- c(
  # Size / institutional structure
  "total_students_enrolled",
  "student_to_staff_ratio",
  "estimated_staff",
  
  # Internationalisation / gender
  "international_students_fraction",
  "total_international_students",
  "female_students_fraction",
  
  # Research output / impact
  "total_citations",
  "citations_per_faculty",
  "total_publications",
  "h_index",
  "x2_year_mean_citations_per_publication",
  "x5_year_publication_growth_percent",
  "x5_year_citation_growth_percent",
  "research_productivity",
  "international_co_authorship",
  "alumni_impact_index_0_1000",
  
  # Geographic / demographic variables
  "country_gdp_billion_usd",
  "country_gdp_per_capita_usd",
  "number_of_associated_institutions",
  "number_of_child_institutions",
  "university_age"
)

outlier_vars <- outlier_vars[outlier_vars %in% names(df_analysis_clean)]

outlier_flags <- df_analysis_clean %>%
  mutate(
    across(
      all_of(outlier_vars),
      ~ detect_outliers_iqr(.x, multiplier = 3),
      .names = "outlier_{.col}"
    )
  ) %>%
  dplyr::select(starts_with("outlier_")) %>%
  mutate(
    n_outlier_flags = rowSums(across(everything()), na.rm = TRUE)
  )

df_outlier_inspection <- df_analysis_clean %>%
  bind_cols(outlier_flags) %>%
  filter(n_outlier_flags > 0) %>%
  dplyr::select(
    institution_name,
    country,
    all_of(outlier_vars),
    starts_with("outlier_"),
    n_outlier_flags
  ) %>%
  arrange(desc(n_outlier_flags))

View(df_outlier_inspection)

df_outlier_inspection %>%
  dplyr::select(
    institution_name,
    country,
    n_outlier_flags
  ) %>%
  arrange(desc(n_outlier_flags)) %>%
  gt()

rows_to_remove <- outlier_flags %>%
  mutate(row_id = row_number()) %>%
  filter(n_outlier_flags > 0) %>%
  pull(row_id)

length(rows_to_remove)

round(length(rows_to_remove) / nrow(df_analysis_clean) * 100, 2)

df_analysis_no_outliers <- df_analysis_clean %>%
  mutate(row_id = row_number()) %>%
  filter(!row_id %in% rows_to_remove) %>%
  dplyr::select(-row_id)

dim(df_analysis_clean)
dim(df_analysis_no_outliers)

df_model <- df_analysis_no_outliers

num_vars_model <- df_model %>%
  dplyr::select(where(is.numeric)) %>%
  names()

cat_vars_model <- df_model %>%
  dplyr::select(where(is.factor)) %>%
  names()

num_vars_model
cat_vars_model
```

### Variable correlations and multicolineality:

```{r}

df_corr <- df_model %>%
  mutate(
    # Log transformations for highly skewed institutional variables
    log_total_students_enrolled = log1p(total_students_enrolled),
    log_estimated_staff = log1p(estimated_staff),
    log_total_international_students = log1p(total_international_students),
    log_total_citations = log1p(total_citations),
    log_citations_per_faculty = log1p(citations_per_faculty),
    log_total_publications = log1p(total_publications),
    log_h_index = log1p(h_index),
    log_research_productivity = log1p(research_productivity),
    log_international_co_authorship = log1p(international_co_authorship),
    
    # Log transformations for geographic / demographic variables
    log_country_gdp_billion_usd = log1p(country_gdp_billion_usd),
    log_country_gdp_per_capita_usd = log1p(country_gdp_per_capita_usd),
    log_number_of_associated_institutions = log1p(number_of_associated_institutions),
    log_number_of_child_institutions = log1p(number_of_child_institutions),
    log_university_age = log1p(university_age)
  )

cor_vars <- c(
  # Size / institutional structure
  "log_total_students_enrolled",
  "student_to_staff_ratio",
  "log_estimated_staff",
  
  # Internationalisation / gender
  "international_students_fraction",
  "log_total_international_students",
  "female_students_fraction",
  
  # Research output / impact
  "log_total_citations",
  "log_citations_per_faculty",
  "log_total_publications",
  "log_h_index",
  "x2_year_mean_citations_per_publication",
  "x5_year_publication_growth_percent",
  "x5_year_citation_growth_percent",
  "log_research_productivity",
  "log_international_co_authorship",
  "alumni_impact_index_0_1000",
  
  # Geographic / demographic variables
  "log_country_gdp_billion_usd",
  "log_country_gdp_per_capita_usd",
  "log_number_of_associated_institutions",
  "log_number_of_child_institutions",
  "log_university_age"
)

cor_vars <- cor_vars[cor_vars %in% names(df_corr)]

short_names <- c(
  "Students",
  "Staff ratio",
  "Staff",
  "Intl frac.",
  "Intl students",
  "Female frac.",
  "Citations",
  "Cit./faculty",
  "Publications",
  "H-index",
  "2Y cit./pub.",
  "5Y pub. growth",
  "5Y cit. growth",
  "Research prod.",
  "Intl co-auth.",
  "Alumni impact",
  "Country GDP",
  "GDP per capita",
  "Assoc. inst.",
  "Child inst.",
  "University age"
)

short_names <- short_names[seq_along(cor_vars)]

cor_matrix <- df_corr %>%
  dplyr::select(all_of(cor_vars)) %>%
  cor(use = "pairwise.complete.obs", method = "spearman")

colnames(cor_matrix) <- short_names
rownames(cor_matrix) <- short_names

corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  order = "hclust",
  tl.cex = 0.65,
  tl.col = "black",
  number.cex = 0.45
)

high_corr_pairs <- as.data.frame(as.table(cor_matrix)) %>%
  dplyr::rename(
    var1 = Var1,
    var2 = Var2,
    correlation = Freq
  ) %>%
  dplyr::filter(var1 != var2) %>%
  mutate(abs_corr = abs(correlation)) %>%
  arrange(desc(abs_corr)) %>%
  distinct(
    pmin(as.character(var1), as.character(var2)),
    pmax(as.character(var1), as.character(var2)),
    .keep_all = TRUE
  ) %>%
  dplyr::filter(abs_corr >= 0.70)

high_corr_pairs %>%
  gt()

ranking_corr_tests <- map_dfr(
  cor_vars,
  \(v) {
    test <- cor.test(
      df_corr[[v]],
      df_corr$world_ranking_the,
      method = "spearman",
      exact = FALSE
    )
    
    tibble(
      variable = v,
      variable_label = get_var_label(v),
      correlation = unname(test$estimate),
      p_value = test$p.value
    )
  }
) %>%
  mutate(
    abs_corr = abs(correlation),
    significant = p_value < 0.05,
    interpretation = case_when(
      correlation < 0 ~ "Higher values associated with better ranking",
      correlation > 0 ~ "Higher values associated with worse ranking",
      TRUE ~ "No monotonic association"
    )
  ) %>%
  arrange(desc(abs_corr))

ranking_corr_tests %>%
  gt()

ggplot(
  ranking_corr_tests,
  aes(
    x = reorder(variable_label, abs_corr),
    y = correlation,
    fill = significant
  )
) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "#0072B2", "FALSE" = "grey70"),
    labels = c("Not significant", "Significant")
  ) +
  labs(
    title = "Correlation of numerical variables with THE ranking position",
    subtitle = "Negative correlations indicate association with better ranking positions",
    x = NULL,
    y = "Spearman correlation",
    fill = "Statistical significance"
  ) +
  theme_minimal()

binary_vars <- c(
  "carbon_neutrality_net_zero_target_0_1",
  "native_english_speaking_country",
  "is_private_university"
)

binary_vars <- binary_vars[binary_vars %in% names(df_corr)]

binary_ranking_summary <- map_dfr(
  binary_vars,
  \(v) {
    df_corr %>%
      group_by(.data[[v]]) %>%
      summarise(
        n = n(),
        mean_ranking = mean(world_ranking_the, na.rm = TRUE),
        median_ranking = median(world_ranking_the, na.rm = TRUE),
        sd_ranking = sd(world_ranking_the, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(
        variable = v,
        variable_label = get_var_label(v),
        category = as.character(.data[[v]])
      ) %>%
      dplyr::select(
        variable,
        variable_label,
        category,
        n,
        mean_ranking,
        median_ranking,
        sd_ranking
      )
  }
)

binary_ranking_summary %>%
  gt()

binary_ranking_tests <- map_dfr(
  binary_vars,
  \(v) {
    
    test_data <- df_corr %>%
      dplyr::select(world_ranking_the, all_of(v)) %>%
      drop_na() %>%
      mutate(
        group = as.factor(.data[[v]])
      )
    
    n_groups <- n_distinct(test_data$group)
    
    if (n_groups != 2) {
      return(
        tibble(
          variable = v,
          variable_label = get_var_label(v),
          test = "Wilcoxon rank-sum test",
          p_value = NA_real_,
          significant = NA,
          note = paste("Skipped: variable has", n_groups, "group(s)")
        )
      )
    }
    
    wilcox_result <- wilcox.test(
      world_ranking_the ~ group,
      data = test_data,
      exact = FALSE
    )
    
    tibble(
      variable = v,
      variable_label = get_var_label(v),
      test = "Wilcoxon rank-sum test",
      p_value = wilcox_result[["p.value"]],
      significant = wilcox_result[["p.value"]] < 0.05,
      note = "OK"
    )
  }
)

binary_ranking_tests %>%
  arrange(p_value) %>%
  gt()

walk(
  binary_vars,
  \(v) {
    p <- ggplot(
      df_corr,
      aes(
        x = .data[[v]],
        y = world_ranking_the,
        fill = .data[[v]]
      )
    ) +
      geom_boxplot(alpha = 0.7) +
      scale_y_reverse() +
      labs(
        title = paste("THE ranking position by", get_var_label(v)),
        subtitle = "Lower values indicate better ranking positions",
        x = get_var_label(v),
        y = "THE ranking position"
      ) +
      theme_minimal() +
      theme(legend.position = "none")
    
    print(p)
  }
)

geo_cat_vars <- c(
  "country",
  "primary_research_topic",
  "top_collaborating_country"
)

geo_cat_vars <- geo_cat_vars[geo_cat_vars %in% names(df_corr)]

geo_cat_ranking_summary <- map_dfr(
  geo_cat_vars,
  \(v) {
    df_corr %>%
      group_by(.data[[v]]) %>%
      summarise(
        n = n(),
        mean_ranking = mean(world_ranking_the, na.rm = TRUE),
        median_ranking = median(world_ranking_the, na.rm = TRUE),
        sd_ranking = sd(world_ranking_the, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      filter(n >= 3) %>%
      arrange(mean_ranking) %>%
      mutate(
        variable = v,
        variable_label = get_var_label(v),
        category = as.character(.data[[v]])
      ) %>%
      dplyr::select(
        variable,
        variable_label,
        category,
        n,
        mean_ranking,
        median_ranking,
        sd_ranking
      )
  }
)

geo_cat_ranking_summary %>%
  gt()
```

### Modeling

```{r}

model_data <- df_corr %>%
  mutate(
    country_grouped = fct_lump_n(
      as.factor(country),
      n = 15,
      other_level = "Other countries"
    ),
    top_collaborating_country_grouped = fct_lump_n(
      as.factor(top_collaborating_country),
      n = 15,
      other_level = "Other collaborating countries"
    ),
    primary_research_topic_grouped = fct_lump_n(
      as.factor(primary_research_topic),
      n = 15,
      other_level = "Other research topics"
    )
  ) %>%
  dplyr::select(
    world_ranking_the,
    
    # Size / institutional structure
    student_to_staff_ratio,
    
    # Internationalisation / gender
    international_students_fraction,
    female_students_fraction,
    
    # Research output / impact
    log_total_publications,
    log_h_index,
    x2_year_mean_citations_per_publication,
    x5_year_publication_growth_percent,
    x5_year_citation_growth_percent,
    log_research_productivity,
    log_international_co_authorship,
    
    # Broader institutional proxies
    alumni_impact_index_0_1000,
    carbon_neutrality_net_zero_target_0_1,
    
    # Geographic / demographic numerical variables
    log_country_gdp_billion_usd,
    log_country_gdp_per_capita_usd,
    log_number_of_associated_institutions,
    log_number_of_child_institutions,
    log_university_age,
    
    # Geographic / demographic binary variables
    native_english_speaking_country,
    is_private_university,
    
    # Geographic / demographic categorical variables
    country_grouped,
    top_collaborating_country_grouped,
    primary_research_topic_grouped
  ) %>%
  drop_na() %>%
  droplevels()

glimpse(model_data)
dim(model_data)

model_data %>%
  dplyr::select(
    country_grouped,
    top_collaborating_country_grouped,
    primary_research_topic_grouped
  ) %>%
  summary()

lm_full <- lm(
  world_ranking_the ~ .,
  data = model_data
)

summary(lm_full)

tidy(lm_full) %>%
  arrange(p.value) %>%
  gt()

glance(lm_full) %>%
  gt()

lm_step_aic <- stepAIC(
  lm_full,
  direction = "both",
  trace = FALSE
)

formula(lm_step_aic)

summary(lm_step_aic)

tidy(lm_step_aic) %>%
  arrange(p.value) %>%
  gt()

glance(lm_step_aic) %>%
  gt()

n_model <- nrow(model_data)

lm_step_bic <- stepAIC(
  lm_full,
  direction = "both",
  k = log(n_model),
  trace = FALSE
)

formula(lm_step_bic)

summary(lm_step_bic)

tidy(lm_step_bic) %>%
  arrange(p.value) %>%
  gt()

glance(lm_step_bic) %>%
  gt()

model_comparison <- bind_rows(
  glance(lm_full) %>% mutate(model = "Full LM"),
  glance(lm_step_aic) %>% mutate(model = "StepAIC LM"),
  glance(lm_step_bic) %>% mutate(model = "StepBIC LM")
) %>%
  dplyr::select(
    model,
    r.squared,
    adj.r.squared,
    sigma,
    AIC,
    BIC,
    p.value
  ) %>%
  arrange(AIC)

model_comparison %>%
  gt()
```

```{r}

set.seed(123)

split_holdout <- initial_split(model_data, prop = 0.8)

data_train <- training(split_holdout)
data_test  <- testing(split_holdout)

set.seed(123)

folds_tbl <- vfold_cv(data_train, v = 10)

m1_main <- lm(
  world_ranking_the ~ .,
  data = data_train
)

summary(m1_main)

tidy(m1_main) %>%
  arrange(p.value) %>%
  gt()

glance(m1_main) %>%
  gt()

full_model <- lm(
  world_ranking_the ~ .,
  data = data_train
)

best_model_aic <- stepAIC(
  full_model,
  direction = "both",
  trace = FALSE
)

formula(best_model_aic)

summary(best_model_aic)

tidy(best_model_aic) %>%
  arrange(p.value) %>%
  gt()

glance(best_model_aic) %>%
  gt()

n_train <- nrow(data_train)

best_model_bic <- stepAIC(
  full_model,
  direction = "both",
  k = log(n_train),
  trace = FALSE
)

formula(best_model_bic)

summary(best_model_bic)

tidy(best_model_bic) %>%
  arrange(p.value) %>%
  gt()

glance(best_model_bic) %>%
  gt()

gam_formula <- world_ranking_the ~
  s(log_h_index) +
  s(log_total_publications) +
  s(log_research_productivity) +
  s(student_to_staff_ratio) +
  s(international_students_fraction) +
  s(log_country_gdp_billion_usd) +
  s(log_country_gdp_per_capita_usd) +
  s(log_university_age) +
  s(log_number_of_associated_institutions) +
  s(log_number_of_child_institutions) +
  female_students_fraction +
  x2_year_mean_citations_per_publication +
  x5_year_publication_growth_percent +
  x5_year_citation_growth_percent +
  log_international_co_authorship +
  alumni_impact_index_0_1000 +
  carbon_neutrality_net_zero_target_0_1 +
  native_english_speaking_country +
  is_private_university +
  country_grouped +
  top_collaborating_country_grouped +
  primary_research_topic_grouped

gam_fit <- gam(
  formula = gam_formula,
  data = data_train,
  method = "REML",
  select = TRUE
)

summary(gam_fit)

compute_reg_metrics <- function(model_fit, data_eval) {
  
  preds <- predict(model_fit, newdata = data_eval)
  
  tibble(
    rmse = sqrt(mean((data_eval$world_ranking_the - preds)^2, na.rm = TRUE)),
    mae = mean(abs(data_eval$world_ranking_the - preds), na.rm = TRUE),
    r_squared = cor(
      data_eval$world_ranking_the,
      preds,
      use = "complete.obs"
    )^2
  )
}

model_tbl <- tibble(
  model = c(
    "M1_main",
    "StepAIC",
    "StepBIC",
    "GAM"
  ),
  fit = list(
    m1_main,
    best_model_aic,
    best_model_bic,
    gam_fit
  )
)

holdout_tbl <- model_tbl %>%
  mutate(
    train = map(fit, ~ compute_reg_metrics(.x, data_train)),
    test = map(fit, ~ compute_reg_metrics(.x, data_test))
  ) %>%
  dplyr::select(model, train, test) %>%
  unnest_wider(train, names_sep = "_") %>%
  unnest_wider(test, names_sep = "_") %>%
  arrange(test_rmse)

holdout_tbl %>%
  gt()

form_m1 <- formula(m1_main)
form_step <- formula(best_model_aic)
form_bic <- formula(best_model_bic)

model_formula_tbl <- tibble(
  model = c("M1_main", "StepAIC", "StepBIC"),
  formula = list(form_m1, form_step, form_bic)
)

cv_one_fold_lm <- function(split_obj, formula) {
  
  data_fold_train <- analysis(split_obj)
  data_fold_test <- assessment(split_obj)
  
  fit <- lm(formula = formula, data = data_fold_train)
  
  compute_reg_metrics(fit, data_fold_test)
}

cv_grid_tbl <- crossing(
  model_formula_tbl,
  folds_tbl
)

cv_long_tbl <- cv_grid_tbl %>%
  mutate(
    metrics = map2(splits, formula, cv_one_fold_lm)
  ) %>%
  dplyr::select(model, id, metrics) %>%
  unnest(metrics)

cv_summary_tbl <- cv_long_tbl %>%
  group_by(model) %>%
  summarise(
    rmse_mean = mean(rmse),
    rmse_sd = sd(rmse),
    mae_mean = mean(mae),
    mae_sd = sd(mae),
    r_squared_mean = mean(r_squared),
    r_squared_sd = sd(r_squared),
    .groups = "drop"
  ) %>%
  arrange(rmse_mean)

cv_summary_tbl %>%
  gt()

cv_one_fold_gam <- function(split_obj, formula) {
  
  data_fold_train <- analysis(split_obj)
  data_fold_test <- assessment(split_obj)
  
  fit <- gam(
    formula = formula,
    data = data_fold_train,
    method = "REML",
    select = TRUE
  )
  
  compute_reg_metrics(fit, data_fold_test)
}

gam_cv_tbl <- folds_tbl %>%
  mutate(
    metrics = map(splits, ~ cv_one_fold_gam(.x, gam_formula))
  ) %>%
  dplyr::select(id, metrics) %>%
  unnest(metrics) %>%
  mutate(model = "GAM")

gam_cv_summary <- gam_cv_tbl %>%
  summarise(
    rmse_mean = mean(rmse),
    rmse_sd = sd(rmse),
    mae_mean = mean(mae),
    mae_sd = sd(mae),
    r_squared_mean = mean(r_squared),
    r_squared_sd = sd(r_squared)
  ) %>%
  mutate(model = "GAM") %>%
  dplyr::select(model, everything())

gam_cv_summary %>%
  gt()

cv_compare_tbl <- bind_rows(
  cv_summary_tbl,
  gam_cv_summary
) %>%
  arrange(rmse_mean)

cv_compare_tbl %>%
  gt()

cv_compare_long <- bind_rows(
  cv_long_tbl,
  gam_cv_tbl
)

ggplot(cv_compare_long, aes(x = model, y = rmse, fill = model)) +
  geom_boxplot(alpha = 0.8, outlier.color = "#D55E00") +
  labs(
    title = "Cross-validation RMSE by model",
    x = NULL,
    y = "RMSE"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

set.seed(123)

data_train_rf <- data_train
data_test_rf  <- data_test

cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)

set.seed(123)

trControl_rf <- trainControl(
  method = "cv",
  number = 10,
  savePredictions = "final",
  allowParallel = TRUE
)

tuneGrid_rf <- expand.grid(
  mtry = c(3, 5, 7, 9, 11),
  splitrule = "variance",
  min.node.size = c(5, 10, 20)
)

rf_fit <- caret::train(
  world_ranking_the ~ .,
  data = data_train_rf,
  method = "ranger",
  metric = "RMSE",
  trControl = trControl_rf,
  tuneGrid = tuneGrid_rf,
  num.trees = 500,
  importance = "impurity"
)

stopCluster(cl)
registerDoSEQ()

rf_fit
plot(rf_fit)

rf_test_pred <- predict(rf_fit, newdata = data_test_rf)

rf_holdout_metrics <- tibble(
  model = "RandomForest",
  test_rmse = sqrt(mean((data_test_rf$world_ranking_the - rf_test_pred)^2)),
  test_mae = mean(abs(data_test_rf$world_ranking_the - rf_test_pred)),
  test_r_squared = cor(data_test_rf$world_ranking_the, rf_test_pred)^2
)

rf_holdout_metrics %>%
  gt()

rf_importance <- varImp(rf_fit)

rf_importance

plot(rf_importance, top = 15)

data_train_nn <- data_train
data_test_nn  <- data_test

dummy_obj <- dummyVars(
  world_ranking_the ~ .,
  data = data_train_nn
)

x_train_nn <- predict(dummy_obj, newdata = data_train_nn) %>%
  as.data.frame()

x_test_nn <- predict(dummy_obj, newdata = data_test_nn) %>%
  as.data.frame()

pp_obj <- preProcess(
  x_train_nn,
  method = c("center", "scale")
)

x_train_nn_scaled <- predict(pp_obj, x_train_nn)
x_test_nn_scaled  <- predict(pp_obj, x_test_nn)

train_nn_final <- bind_cols(
  world_ranking_the = data_train_nn$world_ranking_the,
  x_train_nn_scaled
)

test_nn_final <- bind_cols(
  world_ranking_the = data_test_nn$world_ranking_the,
  x_test_nn_scaled
)

cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)

set.seed(123)

trControl_nn <- trainControl(
  method = "cv",
  number = 10,
  savePredictions = "final",
  allowParallel = TRUE
)

tuneGrid_nn <- expand.grid(
  size = c(3, 5),
  decay = c(0.001, 0.01, 0.1)
)

nn_fit <- caret::train(
  world_ranking_the ~ .,
  data = train_nn_final,
  method = "nnet",
  metric = "RMSE",
  trControl = trControl_nn,
  tuneGrid = tuneGrid_nn,
  trace = FALSE,
  linout = TRUE,
  maxit = 700,
  MaxNWts = 50000
)

stopCluster(cl)
registerDoSEQ()

nn_fit
plot(nn_fit)

nn_test_pred <- predict(nn_fit, newdata = test_nn_final)

nn_holdout_metrics <- tibble(
  model = "NeuralNet",
  test_rmse = sqrt(mean((test_nn_final$world_ranking_the - nn_test_pred)^2)),
  test_mae = mean(abs(test_nn_final$world_ranking_the - nn_test_pred)),
  test_r_squared = cor(test_nn_final$world_ranking_the, nn_test_pred)^2
)

nn_holdout_metrics %>%
  gt()

linear_gam_test_tbl <- holdout_tbl %>%
  transmute(
    model,
    test_rmse,
    test_mae,
    test_r_squared
  )

final_test_compare_tbl <- bind_rows(
  linear_gam_test_tbl,
  rf_holdout_metrics,
  nn_holdout_metrics
) %>%
  arrange(test_rmse)

final_test_compare_tbl %>%
  gt()

rf_cv_summary <- rf_fit$results %>%
  filter(
    mtry == rf_fit$bestTune$mtry,
    min.node.size == rf_fit$bestTune$min.node.size
  ) %>%
  transmute(
    model = "RandomForest",
    rmse_mean = RMSE,
    rmse_sd = RMSESD,
    mae_mean = MAE,
    mae_sd = MAESD,
    r_squared_mean = Rsquared,
    r_squared_sd = RsquaredSD
  )

nn_cv_summary <- nn_fit$results %>%
  filter(
    size == nn_fit$bestTune$size,
    decay == nn_fit$bestTune$decay
  ) %>%
  transmute(
    model = "NeuralNet",
    rmse_mean = RMSE,
    rmse_sd = RMSESD,
    mae_mean = MAE,
    mae_sd = MAESD,
    r_squared_mean = Rsquared,
    r_squared_sd = RsquaredSD
  )

final_cv_compare_tbl <- bind_rows(
  cv_compare_tbl,
  rf_cv_summary,
  nn_cv_summary
) %>%
  arrange(rmse_mean)

final_cv_compare_tbl %>%
  gt()

ggplot(final_test_compare_tbl, aes(x = reorder(model, test_rmse), y = test_rmse, fill = model)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Test RMSE by model",
    x = NULL,
    y = "Test RMSE"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

final_cv_compare_long <- bind_rows(
  cv_compare_long,
  rf_fit$pred %>%
    transmute(
      model = "RandomForest",
      rmse = abs(obs - pred)
    ),
  nn_fit$pred %>%
    transmute(
      model = "NeuralNet",
      rmse = abs(obs - pred)
    )
)

ggplot(final_cv_compare_long, aes(x = model, y = rmse, fill = model)) +
  geom_boxplot(alpha = 0.8, outlier.color = "#D55E00") +
  labs(
    title = "Cross-validation prediction errors by model",
    x = NULL,
    y = "Absolute prediction error"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Conclusions

Random Forest: best model for prediction accuracy.

GAM: best balance between predictive performance and interpretability, capturing non-linear effects without excessive complexity.

Neural Network: not recommended in the expanded dataset; performance deteriorates after adding geographic/demographic variables.

Linear models: useful as interpretable benchmarks, but consistently outperformed by Random Forest and GAM.

Geographic/demographic variables: add contextual information, but do not improve model performance consistently and appear to introduce noise.
